\documentclass[twocolumn, a4paper]{report}

\input{../auxiliary/preamble}

\clearpage
\section*{Acknowledgement}

\begin{abstract}
    I posit that the conceptual architecture we create, i.e. the generative model, obides by the same principles as any hierarchical self-organising system, namely discernment about what it is and what it is not. 
\end{abstract}

\tableofcontents
\listoffigures
\listoftables

Cool words to use: 
\begin{itemize}
    \item explanadum
\end{itemize}



\todo[inline]{Go through text and see which words should be operationalised into the glossary}
\todo[inline]{Put all terms in gpt and ask for similar terms}
\todo[inline]{Use plagiarism checker}
\todo[inline]{make sure that the tone of the thesis is the same everywhere. 
Write it with wit. Make it fun to read. 
Don't make convoluted sentences.}

\todo[inline]{Check for british vs american grammar}

\input{../sections/introduction}

\begin{description}
    \item[computational] Why do things work the way they do?
    What is the goal of the computation?
    What are the unifying principles?
    here we describe the problem as a self-organizing system, building hierarchical generative models. 
    \item[algorithmic] What representations can implement
    such computations? 
    How does the choice of representations
    determine the algorithm? 
    Here we describe the FEP and how the discrimination between me and world gives rise to self-organization
    Here we take the stance of a LOT and talk about the benefits of a symbolic paradigm although this could also be done in other ways, e.g. GLOM
    \item[implementational] here we talk about program synthesis and combinatorial search, MCMC, RL, DreamCoder, etc. 
\end{description}

Other things to clarify
\begin{itemize}
    \item What is a symbol (also relate to semiotics but also in the cognitive sense. Something that relates to something else, a pointer. Think of memories that are stored in the hippocampus.) what about bioelectrical prepatterns? are all representations symbolic? How would symbols be implemented in the brain?
    \item Analytic vs. Geometric Solutions as an analogy to approximation vs symbol? imitating understanding vs actual and think about whether there is actually any difference. perhaps a limitation of the symbolic programming approach is that it is rigid and less flexible?
\end{itemize}


\input{../sections/theoretical.tex}
\newpage
\input{../sections/technical.tex}
\newpage
\input{../sections/discussion}


\Gls{latex} is very useful and can use \gls{glsy}.
\Gls{latex} is very useful and can use \gls{homeostasis}.

\clearpage
\printglossaries

\clearpage
\bibliographystyle{unsrt}
\bibliography{../auxiliary/ref.bib}

\clearpage
\appendix


\end{document}