\section{Introduction}
This thesis aims to delve into the intricacies of cognitive systems, specifically focusing on [model-based and model-free] approaches, and their potential role in the emergence of the 'self'.

In this thesis I will reveal the answer the meaning of life, the universe, and everything.

\info[inline]{structure to keep in mind:}
\begin{itemize}
    \item 3 (or more, see van Gerven) Levels of Marr
    \item Computational
    
    (Building a compositional hierarchical structure)
    \item Algorithmic 
    
(Free energy principle, GFlowNet )
    \item Implementational How can such a system be built in
    hardware? How can neurons carry out the
    computations?
    \item Continuously asking deeper questions as we encounter them. (what is a thought, understanding, etc.)
    \item draw out the shape of the thesis. kind of a v shape?
    \item golden circle (why, what, how)
    \item But \& therefore rule of matt stone and trey parker 
\end{itemize}




\begin{itemize}
    \item First we present the ubiquitous principle of nested multi-scale systems.
    \item We show them in biological systems and how a self could emerge from that. 
    \item We show that the mind also has this structure. 
    \item We hypothesis that imaginaria is structured in the same way. 
\end{itemize}

\subsection{Main Contributions}
The main contributions of this thesis are: 
\begin{itemize}
    \item I show that understanding self-organsiation, moreover the FEP is crucial to understanding the emergence of the self, organisms, cognition.
    \item I hypothesise that the same principle guides our conceptual world. A multi-scale structure that continuiously re-negotiates its concepts. This happens at all levels at the organism. Cells need to discern between friend and foe. A person needs to define concepts (subconscously through this self organisational process.), what is a table and what is not a table. Societies must define concepts such as "what is a woman?". What does marriage mean. What is a democracy. What is Jazz?
    \item A novel method of construcing probabilistic programs 
\end{itemize}